%=========================
% Plantilla LaTeX Artículo Académico
% Compilar con: pdflatex + biber + pdflatex + pdflatex
%=========================
\documentclass[11pt,a4paper]{article}

%-------------------------
% Idioma y tipografía
%-------------------------
\usepackage[english]{babel}
\usepackage{csquotes}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

%-------------------------
% Márgenes y espaciado
%-------------------------
\usepackage[a4paper,margin=2.5cm]{geometry}
\usepackage{setspace}
\onehalfspacing

%-------------------------
% Matemáticas y símbolos
%-------------------------
\usepackage{amsmath,amssymb,amsthm}

%-------------------------
% Tablas y figuras
%-------------------------
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{changepage}

%-------------------------
% Enlaces y PDF
%-------------------------
\usepackage{hyperref}
\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  citecolor=blue,
  urlcolor=blue,
  pdftitle={The Continuum Hypothesis as a speech act},
  pdfauthor={Rodrigo Blanco Betes}
}

%-------------------------
% Listas, código y utilidades
%-------------------------
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{csquotes} % recomendado con biblatex

%-------------------------
% Bibliografía (biblatex + biber)
%-------------------------
\usepackage[
  backend=biber,
  style=apa,
  sortcites=true,
  url=true,
  doi=true
]{biblatex}
\DeclareLanguageMapping{english}{english-apa}
\addbibresource{refs.bib}

%-------------------------
% Metadatos del artículo
%-------------------------
\title{\textbf{The Continuum Hypothesis as a speech act}}

\author{Rodrigo Blanco Betes}


\date{\today}

%-------------------------
% Entornos de teoremas (opcional)
%-------------------------
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}

\begin{document}
\maketitle

%-------------------------
% Resumen y palabras clave
%-------------------------
\begin{abstract}
A recent perspective by \textcite{Feferman2011a} has questioned whether continuum hypothesis (CH) is a well-defined and legitimate mathematical problem. This perspective has received recent criticism by P. Koellner (\citeyear{Koellner2016} \& \citeyear{Koellner2022}) for presenting the risk of reducing mathematics and science to the scope of subjectivism and convention. 

In this text, we approach this debate from a linguistic point of view, drawing on pragmatic theories of meaning and their application to mathematics. We focus particularly on speech act theory, and on the hypothesis—defended by Feferman—that mathematical objects function as elements of social reality, in the sense \textcite{searle1995} applies this concept to domains like politics and interpersonal relations. The purpose of this text is to address whether this point is justified, analyzing if speech act theory and Searle's account of institutions could back up Feferman's perspective, or at least contribute to a self-consistent account of mathematical reality. 
\end{abstract}

\noindent\textbf{Key words:} Continuum Hypothesis, Conceptual Structuralism, Speech Act Theory, Institution Theory.

%=========================
\section{Introduction}

In 1883, Cantor, after extending the concept of cardinality to infinite sets, posed the known as Continuum Hypothesis, stating that there could be no strictly intermediate set between the continuum and the set of the integers. After the development of axiomatic set theory, it was shown to be neither refutable (Gödel, \citeyear{Godel1940}) nor provable (Cohen, \citeyear{Cohen1966}) by the axioms of ZFC set theory, the ones more widely accepted. Later there have been many ways to try to fix the problem, but neither of them has been successful.

The logician and philosopher of mathematics Solomon Feferman proposed a new and controversial way to look at the problem. He developed a philosophical conception of mathematics called conceptual structuralism, which is based on the idea that mathematical notions originate in processes and activities from real experience that help us understand and model our environment. Around certain considerations, he was led to conclude that CH is an intrinsically vague problem, lacking any mathematical meaning.

This view has not been without its detractors. For example, views such as the ones of \textcite{Woodin2001a} or \textcite{Maddy1997} would directly contradict it, for considering the continuum problem to have a definite truth value, and the main goal of set-theorists to settle it through the introduction of new axioms. But Koellner (\citeyear{Koellner2016} \& \citeyear{Koellner2022}) expressed his disagreement in a much more direct way. One of the reasons for this disagreement was an important and peculiar feature of Feferman's conception: its institutional account of the reality and objectivity of mathematical entities. Feferman, in the frame of conceptual structuralism, considers in (\citeyear{Feferman2011a}, p. 10) that mathematical entities' objectivity lies upon their consistent appearance in independently working individuals; i.e. upon an “intersubjective objectivity that is ubiquitous in social reality”. He considers that mathematical entities exist objectively but are dependent on human constructions, following on this regard Searle (\citeyear{searle1995}, p.1): 

\begin{adjustwidth}{1cm}{1cm}
\small
\begin{quote}
[T]here are portions of the real world, objective facts in the world, that are only facts by human agreement. In a sense there are things that exist only because we believe them to exist (…). Things like money, property, governments, and marriages. Yet many facts regarding these things are ‘objective' facts in the sense that they are not a matter of preferences, evaluations, or moral attitudes. 
\end{quote}
\end{adjustwidth}

In any case, he did not explain explicitly the way that mathematical entities could be socially constructed. While in Searle's account we see quite explicitly, both in (\citeyear{searle1995}) and in (\citeyear{searle2010}), the development of a social institution theory in terms of speech act theory, Feferman simply points out that through a certain social conception of mathematics the truth value of determinate mathematical propositions, including CH, could be indeterminate: 

\begin{adjustwidth}{1cm}{1cm}
\small
\begin{quote}
My claim is that the basic conceptions of mathematics and their elaboration are also social constructions and that the objective reality that we ascribe to mathematics is simply the result of intersubjective objectivity (…). That is, it may simply be undecided under a given conception whether a given statement in the language of that conception has a determinate truth value. 
\end{quote}
\hfill \textcite[p. 13]{Feferman2011a}
\end{adjustwidth}

Maybe the lack of explicitness in this conception of mathematics is what led Koellner to criticize it. In his view, there is a huge difference between how political and social institutions come to exist and how numbers or mathematical entities do. For example, he defends (\citeyear{Koellner2022}, p. …) that the reality of the fact that “X is a Spanish citizen” is dependent on institutions such as governments, while the one of “there are infinitely many primes” is not. In general, he considers a huge mistake to make mathematical entities dependent on the intuition or the agreements of a determinate group of people, and that fully undermines, in his view, Feferman's case on CH. 

My main thesis in this text will be that mathematical entities do share, as Feferman points out, many of the features of socially construed entities. In order to defend that, I am going to try to state in precise terms, following in most aspects Searle's theory of social institutions, how the construction of mathematical entities (and, more precisely, domains) can be understood in a similar way to the establishment of social institutions.3 Once this parallelism has been properly addressed, it will follow in a natural way, I hold, that there are some previous preparatory conditions or presuppositions to the acceptance of a determinate mathematical entity. First, the entities introduced would have to be useful in a systematic way to solve, simplify, or generalize previously defined mathematical or scientific questions. For this reason, often the entities introduced should have a certain correspondence with a previously existent pre-theoretical concept, so that their application in those questions can be more straightforward. These conditions are somewhat coincident, and in any case compatible, with Feferman's case that CH is an ill-defined mathematical problem, something I will explore in the end of the article. In general terms, I think that the institutional account for mathematical reality that I am going to explore in this text fits well with many aspects of Feferman's conception of mathematics. 

\textbf{TEXT ORGANIZATION}

%=========================
\section{An institutional account for mathematical structures}

The controversy on the possibility that mathematical entities could have a social or institutional nature is not reduced to conceptual structuralism at all. In fact, there have been many accounts trying to justify that claim (Mention Wittgenstein, Lakatos, Paul Ernest). In recent times, one of the most impactful and explicit of these accounts has been the one of \textcite{Cole2013, Cole2008MathematicalDomains}, who, similarly to Feferman, follows the main Searle's notions on  social institutions. 

Searle's (\citeyear{searle1995} \& \citeyear{searle2010}) central argument is that we create social reality by collectively imposing status functions. According to this notion, institutional facets of reality are developed through a logical process that can be resumed by the formula $\mathrm{X}$ counts as $\mathrm{Y}$ in C, where $\mathrm{X}$is a physical object, $\mathrm{Y}$ is a status function we assign to $\mathrm{X}$ and $\mathrm{C}$ is the relevant context or community. This status function carries a set of deontic powers —rights, duties, obligations and authorizations. For example, the collective acceptance that a piece of paper $\mathrm{X}$ counts as money $\mathrm{Y}$ creates the right for its holder to make payments with it, to pay off a debt, to pay taxes, etc.  

\textcite{Cole2013} argues that Searle's framework of status functions is insufficient for a complete theory of institutions. To account for entities like borders, games, and literary characters, Cole introduces the concept of a representational function. He defines this as an institutional function whose primary purpose is to "facilitate our abilities to represent, analyze, reason about, [and] discover truths concerning... facets of reality that are not the entities in question" (p. 14). While status functions create deontic powers, representational functions equip us with cognitive tools to model and understand specific aspects of reality. For Cole, representational functions conform to a necessary component for explaining a vast range of institutional phenomena.\footnote{See \textcite[14--22]{Cole2013} for more details and some examples.}

This distinction is crucial for our topic. The position of \textcite[1]{Cole2013} is that mathematical structures themselves are “freestanding institutional entities introduced to serve representational functions”. Although his literal phrasing often refers to “mathematical domains” he explicitly clarifies that his underlying commitment is to structures, stating that it is “mathematical structures that exist in virtue of collective agreement”.\footnote{One can see on this regard footnote \textcite[1]{Cole2013}: “More accurately, it is mathematical structures that exist in virtue of collective agreement. Yet, to avoid discussing structuralism, I shall write about mathematical domains”.} I understand such a structure as a specific arrangement of objects and the relations between them, which amounts to endowing a set with additional features—such as operations, topologies, or order relations. The term \textit{freestanding} refers to the fact that these representational functions are created \textit{ex nihilo}, without the need for a preexistent physical object these representational functions are assigned to (the $\mathrm{X}$ in the previous formula). In summary, within Cole's framework, mathematical structures come to exist by being institutionally introduced in the form of representational functions, in such a way that the specific objects that appear in such structures —such as specific set theoretical representations of numbers— do not matter. This conception fits particularly well with the \textit{ante rem} structuralist view of mathematics, for which the specific nature of mathematical objects is non-essential; what is fundamental is the network of relations that constitutes the full structure. 

In Searle's framework, the idea of a social institution is closely related to a form of speech acts called Declarations, which have the unique property of changing the reality to match the propositional content of the speech act simply by successfully performing the act. The logical core of the formula $\mathrm{X}$ counts as $\mathrm{Y}$ in $\mathrm{C}$ is a Declaration. However, the imposition of a status or representational function is not always the result of a literal and explicit speech act. The most evident example is in language: it is an institution that provides both status and representational functions but cannot have been introduced by a set of explicit declarations, for those would presuppose language itself. In fact, Searle (...) considers language as the fundamental institution, which makes it possible that all other institutions come to exist.  

Sometimes we impose SFs and RFs using actual Declarations. The Captain of a soccer team might, for instance, `Declare Tom, you will be Our goalkeeper'. Yet this is not the only way in which Tom can become a Team's goalkeeper; he might simply stand by the relevant goal and, thereby, Cause all relevant parties to start talking and thinking about him as the Goalkeeper. Similarly, occasionally people formally Declare that someone Is their friend, yet typically they just start talking and thinking about that Person differently. As \textcite[10]{searle2010} puts it, on many occasions, `we Just linguistically treat or describe, or refer to, or talk about, or even think About an object in a way that creates a reality by representing that reality as Created'. Henceforth, ``Declaration'' shall refer to all such representational acts rather than merely Declarations in the aforementioned narrow sense. 

In other cases, institutions come to exist by tacit collective acceptance, habitual practice, or an informal agreement; these processes are what \textcite[...]{Cole2013} conceptualizes as standing Declarations. The underlying logical form of the resulting institutional fact remains that of a Declaration, even if it was not initiated by an explicit speech act. This point is radically important for mathematics. The most natural view is that mathematical axioms act as declarations that constitute mathematical reality, a perspective we see in \textcite[...]{SchmidtVenturi2024}. However, Gödel's incompleteness theorems demonstrate that there is no recursive set of first-order axioms that can determine up to isomorphism the most fundamental mathematical structures, such as the natural numbers or the real line.5 Therefore, if one accepts that mathematical structures are institutions in the sense defended here—and that most mathematicians work within the framework of first-order logic—then one must conclude that these structures are established and maintained not by a fixed set of axioms, but by standing Declarations of the mathematical community. 


"The number 3 is part of a system of symbols that we use to create and compute numbers. The existence of the number 3 is dependent on that system in a way that the existence of the moon is not." 

Coherence to mathematical practice: mathematical structures: N, R, C... 

Group theory 

%============================
\section{Recent advances in speech act theory regarding mathematics}

1. Speech act theory has traditionally been applied to social institutions.
2. Recent work has begun applying it to mathematics.
3. Definitions, axioms, postulates, and conjectures have been analyzed as speech acts.
4. However, existing work largely focuses on local linguistic acts.
5. The question of how entire mathematical structures become established remains underdeveloped.

\textcite{RuffinoEtAl2021} focus mainly in the assertive and declarative illocutionary acts appearing in mathematics. In the first case, they focus on what they called properly mathematical assertions, characterized as being stated in scientific and formal contexts with high standards of rigor, and being defined in closed mathematical terms.  

Properly mathematical assertions share many of the properties of ordinary assertives, as for instance their illocutionary point, which is to give a precise description of the (mathematical) world. That determines, as we have seen, their direction of fit. In other cases, however, many of the components of the illocutionary act present peculiarities for the case of mathematical assertives. \textcite[66]{searle1969} states for the speaker to have some evidence of the truth of the propositional content as a preparatory condition of any assertive. In the case of mathematics, this condition seems to be stronger, insofar as a proof of the proposition which is asserted tends to be demanded. There are also special kinds of propositional content conditions, such as the exclusion of any indexical, or the fact that the terms occurring in the propositions must have been mathematically well-defined.

Another point the text makes is trying to analyze the difference between theorems, lemmas, corollaries or similar terms that appear in mathematical communication, which is not explained at the propositional level apparently. The authors hold that these are not just ways of speaking, but that they determine the way the mathematical text and thinking is structured: “Choosing what should count as a theorem (or, how to organize some content into lemmas) is often far from being a merely cosmetic decision. Rather, it amounts to the definition of a hierarchy upon the mathematical text, by which the mathematician attributes more or less importance to different ingredients of the discourse.” Moreover, they defend that different illocutionary points of the statements are the reason why mathematicians tend to group them around these categories.

They then focus on mathematical declaratives, whose analysis is similar to the one of assertives. One interesting aspect is the distinction between stipulative and contentual definitions. The first are just abbreviations of longer expressions (e.g. the definition of a commutative group), while the second are definitions which try to extend a previously given concept. One example of this last kind is the topological definition of continuity, which generalizes the more specific and intuitive notion given in analysis. This last kind of declaration is considered a hybrid illocutionary act, which Searle called Assertive Declaration.20 The properties of declaratives in mathematics then depend on the type of definition we are referring to. Many properties, like the propositional content conditions remain the same as in the case of assertives. Others, like the illocutionary point, get an additional assertive component in the case of mathematical assertive declarations.

%---------Axioms and postulates as speech acts-----------
The role speech acts play in axiomatics in modern mathematics has been studied recently, in by \textcite{SchmidtVenturi2024}. These authors explore in this text the possibility that axioms have both a declarative and assertive illocutionary point, with some directive components in the declarative dimension. The main idea they defend is that every axiom is individually an assertion, but a set of axioms has a declarative role. In some sense they are inspired by the way Hilbert understood axioms: “for Hilbert, axioms are not simply assertions whose truth can be assessed by a comparison with some pre-existent reality, neither are they directions for the construction of mathematical objects, but, similarly to definitions, they play an active role in structuring our image of mathematics.”

The authors develop their interpretation around three main components: the definability of axioms, their declarative and directive roles, and their ability to generate assertions in specific contexts. 

For the first of these aspects, they address in which sense axioms are supposed to bring into existence (to define) mathematical objects. According to their view, language is explicitly separated from mathematical reality to allow axioms to act creatively in constructing new theoretical perspectives. This separation does not imply commitment to a specific mathematical ontology; the analysis remains neutral with respect to whether mathematics is a fictitious construct or the study of independent objects. This way, axioms create a new “layer” of mathematical reality by establishing a conceptual framework. This framework redefines how existing mathematical facts are interpreted: axioms shape, in a linguistic way, the social reality of mathematical objects.23 

Also following Searle's (\citeyear{searle1969}) theory, axioms have both a constitutive and regulative role in building this mathematical reality. According to the Schmidt and Venturi, individually, axioms set conditions on how mathematical objects must behave in order to belong to the framework defined by the axiomatic set (regulative role). This regulation arises from first-order logic, which expresses facts about mathematical objects but does not imply ontological commitments. On the other hand, collective axioms are constitutive, establishing a framework for the theory. Axioms have a declarative role, defining how objects should be within the mathematical context they construct. The authors argue that, by declaring a system of axioms, one does not bring a set of objects into existence, but rather defines how a set of entities should be structured.25 For example, in Hilbert's approach, the consistency of an axiomatic system implies that there is a domain that satisfies those axioms, but this existence is independent of the declarative act.  

From a directive point of view individual axioms act as permissives, setting conditions under which a valid mathematical statement can be made within the framework defined by the axioms. Like rules in chess, axioms not only define a framework but also provide explicit permissions to operate within that framework. Another key point is the assertability of axioms, which depends on providing them with an interpretation in a specific model. An axiom can only be considered as an assertion once it is interpreted in a model; prior to this, it acts only as a permission of assertability. For example, an axiomatic system about a relation R can be interpreted in terms of geometric congruence or numerical equivalence. In both cases, the axioms take on different meanings based on the model.26 This ability of axioms to allow multiple interpretations is also a key feature of Hilbert's model-theoretic approach. 

Even though some linguistic and institutional aspects determine how mathematical objects are, that does not mean that mathematics is fundamentally a social discipline, like Schmidt and Venturi also seem to notice: “We do not consider any social dimension of the linguistic constructions performed by axioms. As a matter of fact, when a mathematical definition is felicitous and well accomplished, it does not need a collective agreement for its validation (…). Notice that we are not claiming that mathematics does not have a social dimension and does not produce institutions, as any other human linguistic activity. We are only claiming that at the level of definitions, and thus at the level of a definitional account of axioms, this social component is absent, since a mathematician is free in the constitution of a new linguistic item”. 

In this sense they are opposed by Cole (2013), who defends that mathematics can be understood as an institutional reality. He argues that mathematical domains (i.e. structured set of mathematical objects, concepts, and their associated relationships) are freestanding institutional entities and explains the objectivity, necessity, and atemporality of mathematics within this framework. 



%============================
\section{Preparatory conditions: is there a notion of \textit{mathematical authority}?}

The role of speech acts in modern mathematical axiomatics has been recently examined by \textcite{SchmidtVenturi2024}. A central thesis they defend is that a set of axioms serves as a declarative function. On this view, axioms are understood as bringing mathematical objects into existence—that is, as defining them. These axiomatic collections thus play a constitutive role, establishing the framework for a theory by acting as permissives that govern what can be legitimately asserted within a specific model. However, as I have argued in Section 2, mathematical structures are, in general, established prior to any specific set of axioms they may model. The intended structure—such as that of the natural numbers—guides the search for adequate axiomatizations, not the other way around. Therefore, it is not the axioms themselves, but a more fundamental standing Declaration—the collective, ongoing practice of mathematicians to identify and work within that structure—that ultimately defines the mathematical objects in question. In a similar vein, \textcite{RuffinoEtAl2021} offer one of the most systematic recent analyses of speech acts in mathematics. A central focus of their work is on mathematical declaratives, particularly definitions. For instance, they identify (p. 10084) consistency with prior definitions as a preparatory condition for successfully introducing a new one. Their analysis extends to many other dimensions of mathematical speech acts as they were defined by \textcite{SearleVanderveken1985}. 

However, these texts—and the literature at large—do not adequately analyze the preparatory conditions for axioms in their constitutive role, nor do they provide a clear account of such conditions for the standing Declarations that establish fundamental mathematical structures. This gap is crucial to address, as preparatory conditions are often the most determinative dimension of the declarative acts that ground our institutions. For example, most institutional declarations require the speaker to hold a special, recognized authority or to follow a specific procedure in the right circumstances. It seems immediately strange, however, to demand similar requirements for the introduction of a mathematical structure. The preparatory conditions for a standing Declaration that successfully institutes a new mathematical reality cannot be social or hierarchical in the usual sense. Yet, could there be a different sense of mathematical authority that serves as a preparatory condition for a felicitous institution of a mathematical structure?  

In general, authority is not an inherent quality but a conferred status; it arises when an individual's acts or pronouncements are consistently accepted as legitimate by a community, thereby constituting an institution. Furthermore, institutional authority is not just based on individual agents but also on the very procedures, roles, and instruments they employ. The legitimacy of an outcome often depends on its generation through an authoritative process. For example, a verdict delivered by a proper trial has much more authority than it would have if decided by a coin toss. Similarly, a proof in mathematics according to the rules of logic has authority, much more than a conclusion reached by a guess, even if it is compelling, or if an eminent mathematician has made it. In this sense, the means themselves can be seen as institutional agents, operating according to standing declarations that grant them specific powers and command collective acceptance. However, applying this framework immediately raises a deeper question: what specific criteria does legitimize the introduction of a new mathematical structure? There is no unanimously accepted view on what constitutes a legitimate, or felicitous, institutional declaration in mathematics. 

A clue can be given by the highly influential mathematician and philosopher of science Henri Poincaré, who can be considered a historical precursor to the institutional view of mathematics. Poincaré argued that mathematical entities are introduced by conventions. However, these are not arbitrary stipulations; they are guided by a principle of convenience and utility, making them fruitful for scientific and mathematical reasoning. He captures this idea eloquently: 

\begin{adjustwidth}{1cm}{1cm}
\begin{quote}
\small
Conventions are found above all in mathematics... It is precisely from this that they draw their rigor; our mind can affirm because it decrees. Are these decrees arbitrary? No, for if they were, they would be sterile. Our decrees are thus like those of an absolute, but wise, prince who would consult his Council of State. \parencite[3]{Poincare1952}
\end{quote}
\end{adjustwidth}

Following the terminology we are developing, Poincaré's decrees are institutional Declarations, while the “Council of State” represents the corresponding notion mathematical authority, which is a quite elucidatory metaphor; it represents the preparatory conditions of the Declarations introducing mathematical structures. In Poincaré's view (\citeyear{Poincare1952}, Chap. I), the fundamental convention of mathematics—one that is neither derivable from experience nor reducible to pure logic—is the acceptance of the principle of induction in arithmetic. This is essentially accepting that there is a foundational institutional Declaration that constitutes the very structure of the natural numbers. The “wisdom” of accepting this convention lies in its unparalleled fruitfulness for science as a whole. As Poincaré argues, deductive and analytical procedures merely rearrange existing knowledge, while induction is a unique engine of genuine discovery: 

\begin{adjustwidth}{1cm}{1cm}
\begin{quote}
\small
The analytical procedure of construction does not force us to descend, but it leaves us at the same level. We can only ascend by mathematical induction, which alone can teach us something new. Without the aid of this induction (...), construction would be powerless to create science. \parencite[28]{Poincare1952}
\end{quote}
\end{adjustwidth}

Poincaré's conventionalism had clear boundaries. He was profoundly reluctant to admit highly abstract entities and structures, such as those found in Cantorian set theory. His position made him a precursor of predicativism, arguing that only mathematical objects that can be predicatively defined—constructed step-by-step from the natural numbers without circular, self-referencing definitions—are legitimate.6  

As the main tittle of \textit{La science et la hypothése} suggests, mathematicians work with hypotheses. These hypotheses, according to Poincaré, are \textit{conventions}. Thus, these conventions to function rightly as hypotheses, must attain to two questions, (1) Why they are true, (2) What are they useful for.

Fruitfulness in mathematics cannot always be seen \textit{a-priori}, for it may depend on the obtained results. For example, the complex plane was accepted after some formulae were solved or the fundamental theorem of algebra solved.

For this reason, whether the introduction of a layer of mathematical reality is accepted could have to be decided \textit{a-posteriori}, so the process of mathematical authority is dynamic and iterative.

%=========================
\section{Mathematical conjectures and CH}

Our account has considered so far mathematical axioms or structures as institutions that are introduced as representational functions.  

\textcite[10087]{RuffinoEtAl2021} mention the possibility of including conjectures in their study: “Maybe conjectures are, in part, the expression of a commitment to find a proof in the future and, in that sense, commissive acts are after all present in mathematics as well, contrary to what we said at the outset. This, too, requires a deeper discussion”.  

This is not, however, the way we would like to face the continuum problem. We are not holding that it could not be treated that way, but that this would not lead us to the resolution of the question we are facing here. The key point here is the set of axioms, postulates, and mathematical objects that make the CH problem definable. For example, to define it in the ZF set theory one needs to use AC and the Power Set Axiom; this last is, in fact, the one that Feferman puts more doubts on. Therefore, we need to look at the progress that has been made around these kinds of mathematical statements: axioms and postulates.

Una definición matemática es una declaración con la presuposición de una correspondencia con una noción pre-teórica relevante (estructuralismo conceptual) 

Axioms as constitutive rules 

CH (concept of an arbitrary subset of the continuum) as an institutional fact 

Preparatory conditions for that fact 

Does CH fulfill them? … 

\section{On the background \textit{conventions} of mathematics}

The natural number structure

The axioms of set theory

Distinction of structures vs functions

Different approaches: Feferman's clarity, Maddy's fruitfulness, Quine's indispensability, quasi-empiricism, intuitionism.

These different approaches assume different preparatory conditions for the institution of mathematical reality. 

\begin{quote}
Those familiar with mathematical practices might be wary of applying Occam's razor to mathematical domains. Mathematics is not governed by Occam's razor. Rather, it is an underlying methodological feature of many mathematical practices that mathematicians should seek maximal generality. This feature of those practices, particularly in foundational areas such as set theory and category theory, result in the characterization of ever larger mathematical domains. There is no problem here, however, because my application of Occam's razor is not internal to some mathematical practice, but rather takes place within the practice of naturalistic metaphysics (i.e., metaphysics guided by the methodological practices of natural science). Occam's razor is a legitimate tool within this practice, because it is a legitimate tool within the non-mathematical aspects of natural science.\parencite[128]{Cole2008MathematicalDomains}
\end{quote}

%--------Our choice of conventions related to Feferman's case---------
Which is the reality? Perhaps quasi-empirical structuralism.

How could CS be complemented with this idea.

In this case, the accepted institutions would be clear ones and conceptually grounded. Thus, no conception of the cumulative hierarchy is accepted.

In my view, this approach to mathematics bears upong a distinction between adopting a structure and a mathematical set of axioms.

In the case of arithmetic, the conventions are the structure. In the case of set theory, they are the set of axioms.

Thus, the background conventions for mathematical practice do not determine $\mathrm{CH}$, which is something stronger than stating that proof methods can not solve it.

%=========================
\section{Criticism to the institutional conception of mathematics}

being able to represent a state of affairs indirectly by way of an entity with features when that state of affairs cannot be directly so represented can be tremendously helpful when it comes to analyzing, reasoning about, discovering truths concerning, etc. the state of affairs in question \textcite[12]{Cole2013}

A social constructivist should maintain that all there is to a particular mathematical domain existing is the undertaking of a mathematical practice that centers about a (coherent) characterization of the structure of the domain in question. \textcite[117]{Cole2008MathematicalDomains}


%--------Contingency of the existence of mathematical objects----------------
Perhaps, when we get to the fringes of contemporary mathematics, we might be willing to grant this kind of contingency thesis. But are we willing to grant that the existence of the natural number structure is a contingent matter? That things really could have gone another way, and we could have failed to construct that structure? This consequence of social constructivism is not quite so easy to accept. But a social constructivist about mathematics must accept it. \textcite[323]{Dieterle2010}

Wittgenstein: en un lenguaje perfecto las matemáticas no existirían (Tractatus) 

Si, dadas las cualidades de la inteligencia humana, los números naturales tienen que existir para que podamos comprender el mundo, ¿esto implica que los números naturales son necesarios, en el sentido lógico? 

Cole's response: atemporality and amodality

%----------The problem of objectivity-----------
Although he does not say so explicitly, I take it that he so limits the discussion because, if such constitutive judgments are the products of mathematicians' rational activities, they cannot be metaphysically objective. Further, they cannot inherit their objectivity from logic, since they are not logical consequences of anything. \textcite[325]{Dieterle2010}

Koellner's point

Paralelismo entre los objetos matemáticos como construcciones y los derechos humanos como construcciones. Diferencia importante: las afirmaciones matemáticas podrían no tener sentido pero nunca pueden ser falsas. 

%-----------Misc---------
Is the subject matter of mathematics social? (Paul Ernst; Ruben Hersh) 

Debate: 

Los números son instituciones (a veces intrapersonales); las condiciones preparatorias permiten que una sola persona pueda evaluar la autoridad matemática 

La noción de estructura matemática es primitiva 

Contar historia inventada de Peano (creador de los numerous naturales; ya sé que no lo es XD) 

My account allows for the possibility of an isolated individual's introducing a new mathematical domain provided that others would, were they to become aware of it, recognize the existential DPs that it carries. \textcite[27]{Cole2013}

\begin{quote}
A similar challenge has no force against a PDRist, because, according to her, mathematical practices are responsible for mathematicians' (and non-mathematicians') basic pure mathematical beliefs being true. The Peano Axioms are true because they have been accepted as an optimal characterization of a collection of objects, i.e., the natural numbers, appropriate for mathematicians' purposes. More generally, because mathematicians get to decide which mathematical objects should be constituted to serve their purposes, and get to decide which basic claims best characterize such objects, roughly speaking, mathematicians do indeed stipulate the basic truths about mathematical domains. Further, mathematical practices, as spatio-temporally instantiated activities, can causally influence human beings to become (at least minimally) competent participants in them. Consequently, mathematical practices can causally influence human beings to have systematically true pure mathematical beliefs. Think, for example, of how school teachers influence their pupils to become minimally competent participants in mathematical practices. This influence begins with such rudimentary lessons as how to add together two natural numbers, includes an introduction to axiomatic characterization and deduction from axioms, usually in the form of Euclidean Geometry, and will, in the mathematically sophisticated classroom, incorporate discussions of how to characterize the continuity of real valued functions using epsilons and deltas. \parencite[124]{Cole2008MathematicalDomains}
\end{quote}

\begin{quote}
According to social constructivists, a similar situation arises in mathematics. Frequently, one mathematician is responsible for the mathematical community taking an interest in a particular mathematical domain.\textsuperscript{18} Consequently, from the perspective of a social constructivist, one individual is responsible for introducing the mathematical practice that constitutes that domain. Yet mathematical domains are sharable because mathematicians use shared logical tools (e.g., first and higher-order quantification) to characterize and constitute those domains.\textsuperscript{19} These shared logical tools allow mathematicians to characterize the domains they seek to theorize about, specifically, how they are structured into objects, properties and relations. For example, characterizing the structure of the domain of natural numbers involves characterizing an $\omega$-sequence. \parencite[128]{Cole2008MathematicalDomains}
\end{quote}

%=========================
\section{Conclusion}
What is to expect then from CH regarding this account? I don't think this is a question to be answered right now, but we can certainly approach it. First, the structures where CH can be formulated in have to be introduced: for example, the domain of third order arithmetic. Second, there is an evaluation of how this structure has been introduced, where a sense of \textit{mathematical authority} comes into play. Then, a the mathematical audience may judge, and decide whether this introduction is metodologically acceptable or not. Thus, a preparatory condition for this declaration, the one where the domain of third-order arithmetic was introduced, could come into play. 

\begin{quote}
\textbf{Fact 1:} Mathematical objects are created by humans. Not arbitrarily, but from activity with existing mathematical objects, and from the needs of science and daily life.

\textbf{Fact 2:} Once created, mathematical objects can have properties that are difficult for us to discover.
\par\hfill\parencite[p.~16]{Hersh1997}

\textbf{4.} Mathematical objects are a distinct variety of social-historical objects. They're a special part of culture.
\par\hfill\parencite[p.~22]{Hersh1997}
\end{quote}

\printbibliography

\end{document}
