%=========================
% Plantilla LaTeX Artículo Académico
% Compilar con: pdflatex + biber + pdflatex + pdflatex
%=========================
\documentclass[11pt,a4paper]{article}

%-------------------------
% Idioma y tipografía
%-------------------------
\usepackage[english]{babel}
\usepackage{csquotes}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

%-------------------------
% Márgenes y espaciado
%-------------------------
\usepackage[a4paper,margin=2.5cm]{geometry}
\usepackage{setspace}
\onehalfspacing

%-------------------------
% Matemáticas y símbolos
%-------------------------
\usepackage{amsmath,amssymb,amsthm}

%-------------------------
% Tablas y figuras
%-------------------------
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{changepage}

%-------------------------
% Enlaces y PDF
%-------------------------
\usepackage{hyperref}
\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  citecolor=blue,
  urlcolor=blue,
  pdftitle={Feferman revisited: conceptual structuralism, Koellner's critique and mathematical practice},
  pdfauthor={Rodrigo Blanco Betes}
}

%-------------------------
% Listas, código y utilidades
%-------------------------
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{csquotes} % recomendado con biblatex

%-------------------------
% Bibliografía (biblatex + biber)
%-------------------------
\usepackage[
  backend=biber,
  style=apa,
  sortcites=true,
  url=true,
  doi=true
]{biblatex}
\DeclareLanguageMapping{english}{english-apa}
\addbibresource{refs.bib}

%-------------------------
% Metadatos del artículo
%-------------------------
\title{\textbf{Feferman revisited: conceptual structuralism, Koellner's critique and mathematical practice}}

\author{Rodrigo Blanco Betes}


\date{\today}

%-------------------------
% Entornos de teoremas
%-------------------------
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem*{theorem*}{Theorem}
\newtheorem{lemma}{Lemma}

\begin{document}
\maketitle

\begin{abstract}
Solomon Feferman \textcite{Feferman2011a} argued that the Continuum Hypothesis may fail to constitute a definite mathematical question. More recently, \textcite{Koellner2016,Koellner2022} has criticized this position as obscure, unfounded, and overly conventionalist. In this paper, I revisit Feferman's perspective through the lens of mathematical practice. I argue that mathematical definiteness cannot be secured purely through categoricity arguments or metamathematical considerations; issues concerning applicability and mathematical practice must also be taken into account, though not in a simplistic indispensability-theoretic sense. Interpreted in this way, Feferman's position can withstand some of Koellner's strongest objections.
\end{abstract}

\section{Introduction}

At the end of the 19th century, Georg Cantor extended the concept of cardinality to infinite sets. In doing so, he was able to show that the cardinality of the continuum is strictly greater than the one of the set of all natural numbers. This led him to conjecture that there was no third set strictly between those two, in terms of cardinality; a conjecture known as the Continuum Hypothesis (CH). The same process can be extended to any set $A$, which can be shown to have smaller cardinality than its power set $\mathcal{P}(A)$. One can then formulate what is known as the Generalized Continuum Hypothesis (GCH), which states that there cannot be an intermediate set strictly between these two in terms of cardinality.\footnote{This intermediate set, in case of existing, would be called an interpolant. There is a stronger way of formulating this hypothesis in terms of well-ordered sets, but it presumes the axiom of choice; see \textcite{Koellner2023}. Even using the weaker version, CH presupposes a huge set theoretical apparatus, consisting in a system strong enough to entail the existence of a definite set containing all the subsets of the continuum, or a strong way of the power set axiom in the case of GCH. In this paper, I focus on the the weaker version, which is also the main focus of Feferman. }

Cantor was unable to resolve this problem, and its status remained open for decades. Years later, \textcite{Godel1940} proved that GCH is consistent with the axioms of ZFC of set theory, and \textcite{Cohen1966} proved the same for its negation. All this gave CH its highly controversial status: an apparently well-defined and legitimate mathematical question but proven to be unsolvable with the tools available to mathematicians in everyday practice.

Since then, there have been a huge number of attempts to settle CH, but none of them has been successful.\footnote{For a summary of the current status of CH, see \textcite{Feferman2011a}, Section 1. } These attempts have never offered a clear counterexample to Cantor's hypothesis, nor an intuitive model demonstrating whether it is true or false. It seems that there is no way to relate CH to a well-established and accepted result of mathematics; so it stands as a metaphysical statement, independent of all the results accepted in everyday mathematical practice.

This has led some mathematicians and philosophers of mathematics to doubt whether it even makes sense to seek a unique and meaningful answer to the continuum hypothesis. One of the most important proponents of this perspective is Solomon Feferman, who proposed a philosophical framework for understanding mathematics called conceptual structuralism. In this framework, mathematics is the result of open-ended processes derived from simple real-life situations, such as counting and ordering, where clarity and practical applicability are crucial. \textcite{Feferman2011a} defended that if we accept conceptual structuralism, then CH turns into an intrinsically vague and indefinite mathematical question. However, conceptual structuralism and its implications for CH have been subject to substantial criticism. Some of the strongest and most direct criticisms have come from Peter Koellner (\citeyear{Koellner2016,Koellner2022}). Koellner's viewpoint on CH bears similarities to those of Woodin (\citeyear{Woodin2001a,Woodin2001b}) or \textcite{Maddy1997}, but he distinguishes himself by having addressed Feferman's philosophical proposal more explicitly.\footnote{\textcite{Parsons2016b} also argues directly against Feferman's conception of set theory, but Koellner's does address Feferman's case on CH much more explicitly.} Shortly put, Koellner's main claims are:

\begin{enumerate}
\item That conceptual structuralism is ambiguous on the status of CH.
\item That Feferman's case relies on a notion of mathematical clarity with several weaknesses, such as the fact that the metamathematical characteristics of set theory that justify Feferman's position are also present in other areas, such as arithmetic.
\item That it is not acceptable to make mathematics dependent on subjective processes and social conventions as Feferman proposes.
\end{enumerate}

In this paper, I offer an interpretation of conceptual structuralism that makes it easier to understand why it leads to the indefiniteness of CH. I defend that once we consider the emphasis conceptual structuralism places on practical applicability—not much treated in Koellner's criticism, not even in Feferman's reasoning—then an argument could be made for the indefiniteness of CH. I do not claim, however, that this argument is definitive, but rather that it is not true that conceptual structuralism is irrelevant to the status of CH. This argument is also, in my view, more promising than Feferman's idea of clarity. Thus, regarding point (2), I lean more towards Koellner's view.

A proper treatment of point (3) above would require a deep examination of speech act theory and institutional theory; something that is well beyond the scope of this paper. Let me briefly point out, however, that I do not think, as Koellner does, that it is impossible to understand mathematical entities in an institutional or social way. Quite the contrary. There have been, for the time being, promising attempts to understand mathematics in an institutional way, as Feferman does, and they have proven to be quite resilient to criticism.\footnote{Take for example \textcite{Cole2013}, who gives an institutional account of mathematics based on speech act theory and argues that it can give an account of the necessity and objectivity of mathematical facts.}

In summary, the most important thesis this paper defends is that an alternative interpretation can be given to conceptual structuralism, which focuses most on the applicability of mathematical structures and conceptions in real-life practice, and less on logical and epistemological clarity. 

The text is organized as follows. In the next section, I interpret Feferman's conceptual structuralism, clarifying and completing it, while also focusing on its implications for CH. Section 3 explains how its relation to practical and scientific applicability could support Feferman's case, despite presenting limitations. In Section 4, I show how this argument applies to actual mathematical practice. In Sections 5 and 6, I address different ways Feferman approaches the notion of clarity and evaluate the problems his arguments present. Finally, in Section 7 I approach Feferman's formal argument, and see how it relates to the philosophical ones analyzed in the rest of the text.

\section{Understanding Feferman's conceptual structuralism}
\label{conceptual-structuralism-intro}

It is not always easy to follow the essential claims of Feferman's conceptual structuralism.\footnote{Other approaches to the philosophy of mathematics have also been characterized as \textit{conceptual structuralism}, such as that of \textcite[postnote]{Isaacson2008}. Unless otherwise stated, however, references to conceptual structuralism in this text concern Feferman's version of the view.} The fundamental idea of this conception is that mathematical notions come from structures based on processes of real experience of activities such as counting, ordering, joining, combining, etc.; and that it is within these structures that mathematical objects and systems are derived and constructed. However, Feferman \citeyear{Feferman2011a} characterizes it in ten theses that are arguably overly concise. He himself acknowledged that they would require further elaboration, but he did not have the space to provide it:

\begin{adjustwidth}{1cm}{1cm}
\begin{quote}
\small
There is no time to elaborate these points, but various aspects of them will come up in our discussion of two constellations of structural notions, first of objects generated by one or more “successor” operations, and second of the continuum. \textcite[p. 12]{Feferman2011a}.
\end{quote}
\end{adjustwidth}

This concerns us here because the most direct criticism by Koellner of Feferman's argument is precisely that his philosophical conception does not say anything about whether CH must be considered a legitimate or definite statement:

\begin{adjustwidth}{1cm}{1cm}
\begin{quote}
\small
The main point I wish to stress is that conceptual structuralism does not on its own say anything about whether or not a given conception (the natural numbers, arbitrary sets of natural numbers) leads to definite or indefinite statements. We are thus provided with a flexible framework, one that is capable of being applied to views on which CH is definite and also to views on which CH is indefinite. \textcite[p. 10]{Koellner2022}.
\end{quote}
\end{adjustwidth}

%------------------------------
I think that Feferman's writings on the topic were more successful in making his position explicit than in defending it. As we will see in this section, he argues, both philosophically and formally, that mathematical conceptions must be clear and conceptually and epistemologically grounded. However, he is less explicit about why this must be so, and about why conceptions such as the one of the set-theoretical hierarchy fail to satisfy these requirements. I think this gap could be addressed through a closer evaluation of mathematical practice, a point to which I return in Section \ref{science-applicability}.

A central idea in conceptual structuralism is that mathematical conceptions always refer to mathematical structures and must be grounded in simple and elementary processes such as counting, combining, and joining. A recurring theme in Feferman's work is that these processes should correspond closely to human sensorimotor and cognitive capacities, a point emphasized in Thesis 3 of conceptual structuralism:

\begin{adjustwidth}{1cm}{1cm}
\begin{quote}
\small
The basic conceptions of mathematics are of certain kinds of relatively simple ideal world-pictures which are not of objects in isolation but of structures, i.e. coherently conceived groups of objects interconnected by a few simple relations and operations. They are communicated and understood prior to any axiomatics, indeed prior to any systematic logical development. \textcite[p. 11]{Feferman2011a}.
\end{quote}
\end{adjustwidth}

Feferman is likely thinking here of basic conceptions such as the notion of natural a number, the induction principle, and a pre-theoretical idea of the successor operation. From these, some further logical development may be appropriate.\footnote{In this regard, the idea of unfolding is relevant. Unfolding refers to taking a concept, definition, or system and systematically developing its consequences, rules, or elements in a transparent, step-by-step manner. See \textcite{FefermanStrahm2000}.} These basic conceptions then become part of an ideal world: a coherent conceptual scenario adopted as a framework for doing mathematics. They are “ideal” because they abstract away from physical reality, yet they remain “world pictures” insofar as they closely correspond to basic real-life experiences. \footnote{It is interesting to note that, in formulating this thesis, Feferman refers to “ideal world-pictures”, linking the last two words, whereas when discussing the conception of the set-theoretical hierarchy—see, for instance, \textcite[p.~19]{Feferman2011a}—he speaks of an “ideal-world picture”. In \textcite[p.~19]{Feferman2011b}, however, he writes “ideal world pictures”, without a hyphen, which arguably aligns more closely with the first of these interpretations.} Feferman also states here that a mathematical conception must be a conception of a structure, a view that I adopt throughout the rest of this text. A distinction is sometimes drawn between “monomorphic” structures, which are categorically determined, and “polymorphic” ones, which are not—see \textcite[133]{Ferreiros2023ConStructuralism}. This distinction is closely related to Feferman's notion of clarity, though the two are not necessarily interchangeable, as I explain in \ref{categoricity-second-order-logic}.

Through this notion, Feferman distinguishes between clear and less clear conceptions of mathematical entities: a clear conception allows statements about it to have determinate truth values, whereas a less clear one can give rise to vague or indeterminate statements, such as the Continuum Hypothesis. This argument integrates naturally with the maxims of conceptual structuralism, since Thesis 5 states:

\begin{adjustwidth}{1cm}{1cm}
\begin{quote}
\small
Basic conceptions differ in their degree of clarity. One may speak of what is true in a given conception, but that notion of truth may only be partial. Truth in full is applicable only to completely clear conceptions. \textcite[p. 11]{Feferman2011a}.
\end{quote}
\end{adjustwidth}

Here, Feferman explicitly acknowledges that truth is not absolute for all conceptions---only fully clear ones can support determinate truth claims. CH, as a statement arising from less clear set-theoretic conceptions, exemplifies this indeterminacy.

%-----------------------------------------
This philosophical approach, presented in Section 2 of \parencite{Feferman2011a}, has a formal counterpart in Section 3. There, Feferman introduces a method for determining whether a mathematical statement is well defined, building on his earlier work on semi-constructive theories \parencite{Feferman2010}. He argues that the Continuum Hypothesis may ultimately prove to be an indefinite statement. In these papers, Feferman adopts an intermediate position between classical and intuitionistic logic, using classical logic in contexts he considers appropriate while relying on intuitionistic logic elsewhere.
Again, I take this logical approach to CH not to provide a new argument, but rather to offer an explanation. Feferman essentially reformulates his philosophical position in formal terms, sometimes making more explicit some of its otherwise obscure aspects. However, I do not think that this formal framework provides any additional reason to accept his position for those who are unconvinced by the underlying philosophical arguments.

Feferman develops the formal system $\mathrm{SCS}$ (Semi-Constructive Set Theory), based on an intuitionistic version of the Kripke–Platek (KP) axiomatic system together with several additional principles. Schematically, the system can be represented as:

\begin{equation}
\mathrm{SCS} = \mathrm{IKP} + \mathrm{MP} + \mathrm{AC}
\end{equation}

Here, $\mathrm{IKP}$ denotes the intuitionistic Kripke–Platek system with the Axiom of Infinity, $\mathrm{MP}$ denotes Markov's Principle, and $\mathrm{AC}$ denotes a constructive and weaker version of the Axiom of Choice.\footnote{See \textcite[Section 6]{Feferman2010} for further details regarding the added axioms. The system defined there also includes the bounded omniscience scheme and the law of excluded middle for $\Delta_0$ formulae, but \textcite{Rathjen2016} proved these principles to be redundant.}

According to Feferman, the validity of the Law of Excluded Middle, $A \lor \neg A$, for a given formula $A$ indicates that $A$ is definite; similarly, other classical principles are associated with definiteness. $\mathrm{SCS}$ assumes definiteness for $\Delta_0$ formulas and for first-order arithmetical statements. By contrast, the universe of sets as a whole is regarded as indefinite, which explains why the system generally operates within intuitionistic logic.

Feferman subsequently conjectured that CH is indefinite in this axiomatic system; that is, that one cannot derive $\mathrm{CH} \lor \neg \mathrm{CH}$. He went further still, conjecturing that CH also remains indefinite in the stronger system $\mathrm{SCS} + \mathcal{P}(\mathbb{N})$, which postulates the existence of the power set of the natural numbers. Both conjectures were later confirmed by \textcite{Rathjen2016}.

Rathjen, however, was cautious about drawing philosophical conclusions from the result and emphasized instead its technical character:

\begin{adjustwidth}{1cm}{1cm}
\begin{quote}
\small
The objective of this paper is to prove Feferman's conjecture. In this sense it is a technical paper (...). However, as far as the ongoing discussions of the foundational status of CH are concerned, readers will have to form their own conclusions. \textcite[p. 743]{Rathjen2016}.
\end{quote}
\end{adjustwidth}

A similar attitude is expressed by \textcite[p. 516]{Koellner2016}: “It seems antecedently unsurprising that if one assumes only classical logic for a limited domain—say first-order number theory—then one will not be able to prove that classical logic holds for distinctive statements of a richer domain.” This fits well with the idea that the formal approach serves primarily to articulate more precisely what counts as “clear” within conceptual structuralism, rather than to provide an independent justification for Feferman's position. However, unlike Koellner, I am inclined to think that Feferman himself never intended this formal approach to serve a stronger justificatory role, or at least never claimed that explicitly.

This formal approach also reveals a tendency toward predicativism, particularly in its rejection of the full power set operation on the set of natural numbers. In SCS, real numbers are constructed via Cauchy sequences rather than as arbitrary subsets of $\mathbb{N}$. While this suffices for much of analysis, it avoids commitment to the full set-theoretic conception of the continuum.

Feferman nevertheless acknowledges a more moderate position, according to which $\mathcal{P}(\mathbb{N})$ may be accepted while higher levels of the set-theoretic hierarchy are rejected. Importantly, his formal results concerning CH remain valid even under this more permissive stance, since CH remains indefinite in the stronger system $\mathrm{SCS} + \mathcal{P}(\mathbb{N})$. This also fits well with Feferman's broader philosophical outlook, as can be seen, for instance, in (\citeyear[p.~22]{Feferman2014ConceptualStruc}):

\begin{adjustwidth}{1cm}{1cm}
\begin{quote}
\small
The direct apprehension of these [structures] leads one to speak of truth in a structure in a way that may be accepted uncritically when the structure is such as the integers but \textit{may} be put into question when the conception of the structure is less definite as in the case of the geometrical plane or the continuum, and \textit{should} be put into question when it comes to the universe of sets.
\end{quote}
\end{adjustwidth}


\section{Scientific applicability in evaluating Feferman's case}
\label{science-applicability}

We have seen that Feferman imposes important epistemological limitations on mathematical structures, at least if statements formulated within them are to be regarded as having determinate truth values. These limitations are relatively clear—especially when expressed formally through the systems $\mathrm{SCS}$ and $\mathrm{SCS} + Pow(\omega)$—but it is far less clear how they are to be evaluated and ultimately justified on philosophical grounds.

As I explain later in Sections \ref{categoricity-second-order-logic} and \ref{standard-models-epistemology}, such claims are difficult to justify purely in formal or metamathematical terms. For this reason, I would like to explore a somewhat different approach. Rather than focusing exclusively on formal considerations, this approach asks to what extent the axiomatic systems and structures that would render CH determinate are integrated into scientific and mathematical practice.

This does not mean, of course, that we should naïvely assume that every mathematical structure or theory employed in practice is thereby determinate, nor that all statements formulated within such structures possess determinate truth values. Nevertheless, the integration of a mathematical structure into mathematical practice usually involves at least some degree of commitment to it on the part of practitioners. Full commitment to a structure—that is, a platonistic belief in its reality—would, of course, imply that statements formulated within it are determinate.

For this reason, the nature of scientific and mathematical practice, and the way in which different mathematical structures become integrated into those practices, may play an important role in debates concerning the determinacy of CH. Apart from this, it is also important to understand the extent to which Feferman himself takes mathematical structures to be integrated into practice, beginning with scientific practice outside mathematics. A corresponding discussion of mathematical practice itself is deferred to Section \ref{mathematical-practice}.

Feferman was generally skeptical of the idea that mathematical entities exist in an abstract Platonic realm independent of reality. At the same time, he acknowledged that the laws of physics are formulated in highly idealized terms, and that scientific models do not necessarily correspond directly to the real world:

\begin{adjustwidth}{1cm}{1cm}
\begin{quote}
\small
One must ask whether the mathematical structure of the real number system can be identified with the physical structure, or whether it is instead simply an idealized mathematical model of the latter, much as the laws of physics formulated in mathematical terms are highly idealized models of aspects of physical reality. \textcite[p. 107]{Feferman1999}.
\end{quote}
\end{adjustwidth}

Koellner even regards this as a difficulty for Feferman's position, observing:

\begin{adjustwidth}{1cm}{1cm}
\begin{quote}
\small
There is a hint of an answer to this question in his remarks on the role of mathematical models in physics. For in addition to thinking that no mythical third realm (the platonic realm) can secure the definiteness of set theory, he also thinks that the physical world cannot secure the definiteness of set theory. \textcite[p. 507]{Koellner2016}.
\end{quote}
\end{adjustwidth}

However, the fact that the physical world does not correspond directly to any mathematical structure does not imply that idealized mathematical conceptions are entirely independent of physical reality. What seems most consistent with Feferman's work is instead the view that mathematical entities are justified pragmatically rather than ontologically. On this interpretation, the crucial link between mathematics and reality lies in our practices—in the ways we model, measure, and reason about the world.

This line of thought bears some resemblance to the indispensability arguments of Quine and Putnam, which are noteworthy insofar as they were among the first to emphasize the role of the applicability of mathematics to the world in justifying belief in mathematical entities. Feferman's position, however, differs from theirs in important respects. Whereas Quine and Putnam were primarily concerned with epistemological and ontological questions, Feferman places much greater emphasis on mathematical practice and on the epistemological limitations governing mathematical conceptions.\footnote{Indeed indispensability argument are far from universally accepted among philosophers of mathematics. For a critical discussion, see \textcite{Maddy1992}.}

For this reason, even without assuming a complete correspondence between mathematical entities and physical objects, there remains room to maintain that, within conceptual structuralism, mathematical structures may bear an indirect relation to physical reality through processes of modelling, idealization, and abstraction.

Feferman conjectured that all mathematics applicable to real-world physics is derivable from relatively simple systems such as Peano Arithmetic. If so, systems such as $\mathrm{SCS}$—which reproduce the formal assumptions of conceptual structuralism as an account of mathematics—would suffice to formalize all scientifically relevant mathematical applications without requiring stronger set-theoretic frameworks:

\begin{adjustwidth}{1cm}{1cm}
\begin{quote}
\small
I have come to conjecture that practically all scientifically applicable mathematics can be formalized in systems reducible to PA, or, as I have sloganized it in \textcite{Feferman1993}: a little bit goes a long way. Against this, I have learned of a couple of cases in some approaches to the foundations of quantum field theory where it appears one must go beyond the resources of PA; but the physical theories that require such additional strength are rather speculative. \textcite[p. 15]{Feferman1999}.
\end{quote}
\end{adjustwidth}

This remains, of course, a conjecture, but it is a fairly plausible one. Most physical theories are formulated within the frameworks of special relativity, quantum mechanics, or general relativity, and none of these appears to require especially strong foundational systems. Even if the claim that all such mathematics can be formalized in PA is too strong, stronger but still comparatively weak systems such as $\mathsf{ACA}_0$ remain available.

The system $\mathsf{ACA}_0$ can be formulated within second-order arithmetic and permits the formation of sets of natural numbers definable by arithmetical formulas. Calculus, algebra, functional analysis, basic topology, and many other branches of mathematics used in physics can be formalized within this framework. The stronger system $\Pi^1_1\text{-}\mathsf{CA}_0$, which additionally allows sets of natural numbers defined by $\Pi^1_1$ formulas, is (thought not predicative) generally regarded as more than sufficient. In any case, there is no need to move beyond the language of second-order arithmetic.\footnote{These systems arise in the program of reverse mathematics, which is largely carried out within second-order arithmetic. See \textcite{Simpson2006} for further details.}

By contrast, statements such as the Continuum Hypothesis require systems of third-order arithmetic, quantifying not only over natural numbers and sets of natural numbers (that is, real numbers), but also over sets of sets of natural numbers. This highlights the fact that merely formulating CH already involves a level of set-theoretic complexity far beyond what is required for the core mathematics used in physics. Thus, at least from the standpoint of scientific applicability, there appears to be little reason to regard CH as determinate, given that scientific practice would remain essentially unaffected regardless of whether CH were true or false. Needless to say, scientific applicability is not the only consideration that might support regarding CH as determinate.

This is another point at which Feferman's tendency toward predicativism appears, since he argues \parencite[p.~22]{Feferman2011a} that all current applications of analysis in physics can be carried out predicatively, whithout need of set-theoretically defining the continuum:

\begin{adjustwidth}{1cm}{1cm}
\begin{quote}
\small
What this shows is that the argument from the success of physics to the reality of the set-theoretical conception of the real numbers is completely vitiated.
\end{quote}
\end{adjustwidth}

One might nevertheless argue that this reduces the scientific applicability of mathematical conceptions merely to their capacity to provide the formal tools required for proving relevant results in physics. Scientific and mathematical practice involve more than the development of frameworks within which results can be formulated and proved; they constitute networks of interrelated practices involving, for example, practical applications, communication, pedagogy, and the dissemination of mathematical knowledge.

From this perspective, predicative mathematics may sometimes complicate the construction and development of certain mathematical domains, especially regarding the problem of constructing the continuum in a predicative and non-set-theoretic manner. Still, I think Feferman was right to remain flexible on this issue, as suggested at the end of Section \ref{conceptual-structuralism-intro}. Moreover, his position concerning CH does not appear to be significantly affected here, since the limits of predicative mathematics are generally located within the scope of second-order arithmetic. To assess this issue more fully, however, we must move beyond scientific applicability and consider mathematical practice itself.

\section{Conceptual Structuralism regarding the philosophy of mathematical practice}
\label{mathematical-practice}

For many years, the philosophy of mathematics was dominated by the foundationalist tradition, closely associated with foundational programs within mathematics, such as axiomatic set theory. This situation began to change in the second half of the twentieth century with the emergence of the so-called \textit{philosophy of mathematical practice}. As \textcite{Mancosu2008} observes:

\begin{adjustwidth}{1cm}{1cm}
\begin{quote}
\small
Mathematical logic, which had been essential in the development of the foundationalist programs, was seen as ineffective in dealing with the questions concerning the dynamics of mathematical discovery and the historical development of mathematics itself. \textcite[p. 4]{Mancosu2008}.
\end{quote}
\end{adjustwidth}

Among the central ideas commonly associated with this philosophical tradition are the following:\footnote{In our discussion of mathematical practice, we follow in several respects the way \textcite{Ferreiros2016} approaches the concept; see especially Section 2.3.}

\begin{enumerate}
\item Mathematics is a human activity, not merely a compendium of timeless truths. In this activity, reference to the mathematical community is essential.
\item Explanations, clarity, and heuristics play an important role in mathematics, and not only formal reasoning.
\item Social and historical elements play a crucial role in the development of mathematics. There is not necessarily a unique framework that univocally determines mathematical practice.
\item Theoretical knowledge is essential to mathematics, but must ultimately remain connected to practice.
\end{enumerate}

Regardless of whether one accepts the broader commitments of this philosophical movement, it is easier to agree that any principle consistently guiding mathematical practice constitutes an important datum that the philosophy of mathematics must account for. For this reason, and in addition to the considerations discussed in Section \ref{science-applicability}, it is important to understand how Conceptual Structuralism relates to this perspective, and what role structures in which CH is regarded as determinate play within mathematical practice more generally.

%----------------Maddy's MAXIMICE&UNIFY-----------------------
One of the most influential defenses of methodological principles guiding mathematical practice is that developed by \textcite{Maddy1997}, who emphasizes two complementary maxims: \emph{MAXIMIZE} and \emph{UNIFY}. According to this view, mathematical practice tends to expand the mathematical universe as much as possible, provided consistency is preserved. This resembles the principle of quasi-combinatorialism, that is, the idea that understanding set theory involves grasping the notion of an arbitrary set—definable or not—and therefore accepting as many sets as possible.

Such principles were undoubtedly important in the early development of set theory, but they have not remained uniformly dominant, either prior to the rise of foundationalism or in contemporary mathematical practice. Moreover, maximizing principles may conflict with one another; consider, for example, the incompatibility between the Axiom of Choice and the Axiom of Determinacy. Furthermore, the very notion of an arbitrary set is itself controversial, partly because it cannot be fully captured within formal systems such as ZFC, or even within considerably stronger systems, as argued by \textcite{Ferreiros2011}.

%----------Feferman & mathematical practice-----------------
Feferman’s conceptual structuralism can in fact be understood as an alternative to unrestricted maximizing and unifying principles; see \parencite[p.~9]{Feferman2011a}. He addresses Maddy’s proposal more directly in \textcite[p.~9]{Feferman2000newaxioms}:

\begin{adjustwidth}{1cm}{1cm}
\small
\begin{quote}
In binding itself to mathematical practice, this kind of naturalism is in danger of being unduly transitory. Even if one takes the proposed naturalistic point of view and mathematical practice as exemplified in set theory for granted, there is a crucial question as to what determines the mathematical ends'' for which the most effective means'' are to be sought. And having chosen the ends, in what sense does effectiveness justify the means?
\end{quote}
\end{adjustwidth}

On the same page, Feferman distances himself further from Maddy’s maxims by arguing that the mathematical universe should, in some sense, remain limited, even if no absolute or rigid limiting principle can be imposed:

\begin{adjustwidth}{1cm}{1cm}
\small
\begin{quote}
Old-style infinitesimals, Dirac delta-functions, unrestricted comprehension? If not, what justifies what is to be admitted to mathematics? Once it is agreed that there has to be some sort of justification, intrinsic or extrinsic, then one is in the game of potentially limiting mathematics in some way or other. I don't hew to any sort of absolute principle in favor of limiting mathematics.
\end{quote}
\end{adjustwidth}

In these passages, Feferman suggests that merely following whatever mathematical practitioners happen to do at a given historical moment may lead to philosophically fragile accounts of mathematics. At the same time, he is not claiming that the philosophy of mathematics should ignore mathematical practice altogether. Rather, he implicitly appeals to mathematical practice in arguing against maximizing conceptions of mathematics, noting that entities or principles such as infinitesimals, Dirac delta-functions, and unrestricted comprehension are no longer accepted in the unrestricted forms in which they originally appeared. In my view, the key is instead to identify methodological principles that remain stable across historical periods and across different domains of mathematical practice.

The question, then, is not whether mathematical practice matters, but which features of mathematical practice could justify regarding certain mathematical structures as determinate. In some cases, the role a structure plays within mathematical practice gives us stronger reasons to regard it as determinate than can be provided by metamathematical results alone. In particular, a structure such as the set $\mathbb{N}$ of natural numbers is treated significantly differently in mathematical practice from structures such as the set-theoretic hierarchy. The following features do not by themselves settle questions of determinacy, but they help explain why certain mathematical structures are treated in practice as more fixed and objective than others:

\begin{enumerate}

  \item \textbf{Modes of access in practice:} Our practical interaction with the natural numbers is iterative, sequential, and intuitive. Some of our most fundamental cognitive capacities, especially those related to space and time, appear to mirror this structure. By contrast, set-theoretic conceptions appear quite differently in mathematical practice: they are more dependent on formal axiom systems and less directly connected to intuition. For instance, ordinary mathematical practice almost never directly manipulates objects such as $\mathcal{P}(\mathbb{R})$.\footnote{If we examine the paradigmatic structures that arise across different areas of mathematics, we find that they often remain below the level of the full power set of the continuum. For example, the canonical structure in topology is the real line endowed with its usual topology, which restricts attention to open sets. Similar observations apply in areas such as geometry, analysis, and measure theory.}

  \item \textbf{Degree of framework dependence:} Arithmetic often functions as a fixed background across different foundational programs. Apart from finitists and certain forms of constructivism and intuitionism, most foundational perspectives accept the basic notions of arithmetic and treat arithmetical statements as determinate. By contrast, higher set theory is often highly sensitive to variations in the surrounding foundational assumptions. This is reflected in the large number of competing axiomatic frameworks that have been proposed for set theory, including $\mathrm{ZFC}$, $KP$, intuitionistic set theory, $\mathrm{SCS}$, and many other cases.
  
  \item \textbf{Explanatory indispensability vs. optional enrichment:} The natural numbers seem unavoidable in ordinary mathematics. By contrast, large portions of higher set theory function more like optional enrichments of mathematics. The Continuum Hypothesis is a paradigmatic example: vast areas of mathematics proceed unaffected regardless of its truth value. Similar considerations apply to large cardinal axioms, determinacy principles, forcing axioms, and related set-theoretic extensions.
  
  \item \textbf{Convergence of representations:} The natural numbers appear more determinate in part because an enormous variety of mathematical practices converge on the same structure. By contrast, the set-theoretic hierarchy admits radically different extensions and interpretations, including forcing extensions, inner models, large cardinal hierarchies, determinacy models, and multiverse conceptions.
  
\end{enumerate}

In Section \ref{standard-models-epistemology}, I return to these differences, focusing in particular on how mathematical practice treats the standard models of arithmetic and set theory. In the context of Conceptual Structuralism, it is natural to view these features of mathematical practice as reflecting one of its central ideas: namely, that determinate structures must remain sufficiently connected to our cognitive capacities. I am not claiming that this automatically proves Feferman right, but rather that a sufficiently detailed examination of mathematical practice may have something to contribute to the question of the definiteness of CH.

I now return to metamathematical arguments, which play an important role in Koellner's criticism. I agree with Koellner that such arguments are not sufficient to establish the indefiniteness of CH, but I also think that he assigns them too much weight in his interpretation of Feferman.

\section{Categoricity and second order logic}
\label{categoricity-second-order-logic}

One of the main metamathematical and model-theoretical approaches to determinacy of mathematical structures is the concept of categoricity. Perhaps the strongest reason for regarding a mathematical structure as determinate would be the existence of a categorical axiomatization in first-order logic. However, the Löwenheim--Skolem theorem implies that no recursively enumerable first-order theory with an infinite model can be categorical. Consequently, neither the structure of the natural numbers nor the set-theoretic hierarchy can be regarded as determinate solely on the basis of first-order categoricity.

Some authors, such as \textcite[31]{Isaacson2008} and \textcite[chapter 8]{shapiro1991foundations}, have argued that categoricity results in full second-order logic should be accepted, since categoricity becomes available in that setting. In the case of arithmetic, full second-order Peano arithmetic, $PA_2$, is categorical. In the case of set theory, an initial segment of the universe $V_\kappa$, where $\kappa$ is a strongly inaccesible cardinal, is likewise quasi-categorically characterized by second-order formulations of Zermelo-Fraenkel set theory, $\mathrm{ZFC}_2$. Both \textcite[19]{Feferman2011a} and \textcite[511]{Koellner2016} reject this position, arguing that full second-order semantics presupposes a background notion of ``all subsets,'' thereby involving a commitment of essentially the same kind and strength as the set-theoretic ontology under discussion. 

In this section, I sketch both the categoricity result for arithmetic and the quasi-categoricity result for set theory. I then examine whether, even if these results fail to establish the determinacy of the corresponding structures, they nevertheless receive different treatment within ordinary mathematical practice.

\subsection{The categoricity of \texorpdfstring{$PA_2$}{PA2}}

The categoricity theorem for second-order Peano Arithmetic ($PA_2$) can be formalized within $ZF\!-\!P+\mathrm{Pow}(\omega)$. Under full second-order semantics, second-order variables range over all subsets of the first-order domain. In the case of arithmetic, this amounts to quantification over $\mathcal{P}(\mathbb{N})$. The crucial principle is the second-order induction axiom:

\[
\forall X\Bigl[(0\in X \wedge \forall n\,(n\in X \rightarrow S(n)\in X))
\rightarrow \forall n\,(n\in X)\Bigr].
\]

Once this principle is assumed together with the remaining axioms of $PA_2$, the proof that every model of the theory is isomorphic to the standard model is straightforward.  Be $(\mathbb{N},0,S)$ the standard model of $PA_2$, and suppose that $(M,0_M, S_M)$ is another structure that models this system of axioms. Let $(\mathbb{N},0,S)$ denote the standard model of $PA_2$, and suppose that

\[
(M,0_M,S_M)\models PA_2.
\]

Define a function $f:\mathbb{N}\to M$ recursively by

\[
f(0)=0_M,
\qquad
f(n+1)=S_M(f(n)).
\]

The successor axioms imply that $f$ is injective. To prove surjectivity, define $A=\{x\in M:\exists n\in\mathbb{N}\,(f(n)=x)\}$. By construction, $0_M\in A$. Moreover, if $x\in A$, then for some $n\in\mathbb{N}$ we have $x=f(n)$, and hence $S_M(x) = S_M(f(n)) = f(n+1)$, so $S_M(x)\in A$. Thus $A$ contains $0_M$ and is closed under $S_M$. By the second-order induction axiom, $
A=M$.

Hence every element of $M$ lies in the image of $f$, so $f$ is surjective. Therefore $f$ is a bijection preserving $0$ and successor, and thus an isomorphism. Consequently every full model of $PA_2$ is isomorphic to $(\mathbb{N},0,S)$.

I have highlighted only those aspects of the proof that are relevant to the assumptions on which it depends, rather than presenting its full formal derivation.\footnote{For a rigorous formulation of the proof, see \textcite[82]{shapiro1991foundations}.}

The crucial point, in my view, is that the underlying set-theoretic assumptions are comparatively weak. The strongest assumptions involved are arguably the existence of the power set of the natural numbers together with a weak form of replacement, both of which tend to be accepted ubiquitously throughout ordinary mathematical practice.

\subsection{The quasi-categoricity of ZFC2}

The quasi-categoricity result of $\mathrm{ZFC}_2$ can be formalized and proven within $\mathrm{ZFC}$, a similar setting that we found in arithmetic. We can formalize in this system the idea of being a model of full second-order set theory by the relation $(M,P) \models \mathrm{ZFC}_2$, where $P=\mathcal{P}(M)$ (and similarly for the interpretation of higher order variables), and in such a way that the $\mathrm{ZFC}$ axioms hold when relativized to $M$, where the replacement schema can now be seen as a single axiom:

\[
\forall f\,\forall x\,\exists y\,\forall z\,
\Bigl(
z\in y \leftrightarrow \exists t\in x\,(z=f(t))
\Bigr).
\]

With this setting, the quasi-categoricity result is stated as follows:

\begin{theorem*}[\textcite{Zermelo1930}]
  
\(\kappa\) is strongly inaccessible iff
\[
V_\kappa \models \mathrm{ZFC}_2 .
\]

\end{theorem*}

\begin{proof}

Except replacement, the axioms of $\mathrm{ZFC}$ easily hold in \(V_\alpha\), in case \(\alpha > \omega\) is a limit ordinal. Replacement is the axiom that uses the hypothesis that $\kappa$ is inaccesible: if $x \in \kappa$ and $f: x \to \kappa$, $f(x)= \{ y \subseteq \kappa | \exists z \in x, f(z) = y \}$, then we have $|f(x)| \leq |x| < \kappa$. As this cardinal is inaccesible, it cannot be reached by taking unions or power sets of smaller cardinals, which implies that $f(x) \in V_\kappa$, thus being replacement verifyed relatively to this model.

To proof the converse, we need to see both that $\kappa$ is regular and that, if $\mu < \kappa$, $\mathcal{P}(\mu) < \kappa$. If $\kappa$ were not regular there would be an $\alpha < \kappa$ and a function $f : \alpha \to \kappa$ such that $max_{x\in\alpha}\{ f(x) \} = \kappa$. Then $f \subseteq V_\kappa$, so by second-order Replacement $Im(f) \in V_\kappa$. Then by the axiom of union in $V_\kappa$, $\cup Im(f) = \kappa \in V_\kappa$, a contradiction. For the second part, suppose there would be a \(\mu < \kappa\) such that $2^\mu \geq \kappa$. Then there would be a surjective function $f : \mathcal{P}(\mu)\to \kappa$, But power set axiom implies $\mathcal{P}(\mu) \in V_\kappa$ so by second-order replacement, $Im(f)= f(\mathcal{P}(\mu))=\kappa \in V_\kappa$, a contradiction.\footnote{See \textcite[23]{Kanamori2003} for a modern formulation and proof of the theorem. See also \ref{set-theory appendix} for precise definitions of the set-theoretic notions employed there.}

\end{proof}

Note that, similarly to the case of arithmetic, the proof is straightforward. The theorem then shows that $V_\kappa$ is the only model of set theory of the corresponding size, what leads to the \textit{quasi-categoricity} label. 

\subsection{Conclusion}

Note that, in the end, this is a theorem of first order $\mathrm{ZFC}$. This is in the sense that $\mathrm{ZFC}_1$ proves that for any strongly inaccesible cardinal $\kappa$, $V_\kappa$ is the only model of the corresponding size of $\mathrm{ZFC}_2$.

Internally, $V_\kappa$ are isomorphic (that is, in a given model, $V_\kappa$ is trivially isomorphic to $V_\kappa$). But externally, two models of $\mathrm{ZFC}$ may have different $V_\kappa$.

\textcite{Koellner2010SecondOrderLogic} defends that second-order logic does not have the absoluteness properties to be called a ``logic''.

The circularity is so patent, one does not see that any more can be claimed thereby than that the definiteness of those notions may be accepted as a matter of faith. \textcite[20]{Feferman2011a}

One might seek to restore clarity through second-order logic, which does yield categorical axiomatizations of arithmetic and of set-theoretic initial segments. 

\begin{adjustwidth}{1cm}{1cm}
\begin{quote}
\small
The categoricity of the $\mathbb{N}$-structure can be obtained in weak second-order logic and does not even depend on its impredicativity. Hence its fully determinate nature, so to speak. In practice, this is admitted by mathematicians and logicians who differ in their acceptance of some questionable foundational principles. \textcite[141]{Ferreiros2023ConStructuralism}
\end{quote}
\end{adjustwidth}

The proof in set theory is more circular than in arithmetic. In either case it is not a mean to stablish definiteness in logical terms.

Definiteness cannot be proven in logical terms, because it trascends logic. It is not to be expected that we can prove logically the definiteness of a certain entity without also departing from the one of a similar one.

In this case what we have is the commitment to a specific mathematical reality. From this reality practicioners extract axioms, that lead them to the results that they are interested in. In the case of the categlricity of arithmetic, they commit to the natural number structure and to some features of its subsets. This features can be formalized in various axioms systems, which suffice in this case to prove the result meeded. In the case of set theory the commitments needed are much stronger, and less usual in current mathematical practice.

In any case, it is the commitment to mathematical structures, and the fact that they can be conceived by practicioners, what leads to the axiomatizacion and, in this case, the categoricity result, but never the other way around.

\section{Standard models}
\label{standard-models-epistemology}

Non-categoricity implies the existence of non-standard models of both arithmetic and set theory, where certain statements can have different truth values. In arithmetic, every non-standard model contains (as an initial segment of its underlying domain) an image of the standard natural numbers. Therefore, non-standard models are easy to identify. In spite of this, Koellner argues that the case in set theory is exactly the same, and that non-standard models are easy to identify as well:

\begin{adjustwidth}{1cm}{1cm}
\begin{quote}
\small
But the case in set theory is exactly parallel. Take a case like Cohen's method of forcing where you build a non-standard model (...). This is done within set theory and it is revealed in the first step that one is dealing with a non-standard model since it is a countable object. In the second approach one builds class size models (...) but one builds a Boolean-valued model and, once again, it is immediately revealed that it is non-standard. \textcite[p. 512]{Koellner2016}.
\end{quote}
\end{adjustwidth}

Note how Koellner correctly identifies methods for recognizing a non-standard model of set theory but fails to address how one would positively identify a standard one. This highlights a fundamental disanalogy between arithmetic and set theory. In arithmetic, the standard model, $\mathbb{N}$, is epistemologically privileged. We have a direct pre-theoretical intuition of this structure—it is the very one we use to learn counting. The complete theory of this model serves as a clear (although not recursively computable) benchmark for truth in arithmetic. However, we have no analogous intuition for the entire cumulative set-theoretic hierarchy, where the standard model of set theory is supposed to depend on. Instead, different set-theoretic frameworks propose competing visions of what the universe of sets should be. Most set-theorists would argue that $L$, Gödel's constructible universe, is not a standard model of the set-theoretic universe. However, to argue this way one might need further philosophical assumptions (such as maximizing principles) which seem much more demanding than arguing that $\mathbb{N}$ is the standard model of arithmetic. The standard model $\mathbb{N}$ is grounded in the simple iterative application of the successor operation, which exhaustively generates the arithmetical universe. By contrast, the cumulative set-theoretic hierarchy is built through a much more intricate network of operations—power set, replacement, transfinite recursion—that do not obviously exhaust the universe of sets.

If we accept conceptual structuralism, a basic conception of mathematics should be of a structure, from which one could later extract a set of axioms describing it. For that to succeed, there is need of an epistemologically privileged structure, such as $\mathbb{N}$. From this structure one can extract, for example, a set of axioms such as PA, and start reasoning about them. However, the process is inverted in the case of set theory: one begins from a set of axioms, such as ZF, and then tries to accommodate them in a determinate structure, such as the cumulative hierarchy of sets. But as the structure is no longer epistemologically privileged, there is often no way of deciding fundamental questions in it without relying on controversial philosophical principles. I think that, in fact, Feferman is very close to thinking this when he states that ``the concept of arbitrary set essential to its formulation is vague or underdetermined and there is no way to sharpen it without violating what it is supposed to be about'' (\textcite{Feferman2011a}, p. 1).

This provides a starting point to address the question of how set theory might be considered unclear while arithmetic is not. However, applying this to CH would be problematic because the definiteness of this statement would only correspond to considering third-order arithmetic as clear. It is more difficult to argue that we cannot depart from an intuitive structure, such as $\mathbb{N}$, and develop the corresponding axioms. It is worth recalling that incongruencies such as independence results appear at all levels: arithmetic, finite-order arithmetic, and set theory. \footnote{Koellner treats this in great detail; see (\citeyear{Koellner2016}), sections 5 and 6.}

\textcite{Feferman2000rationale} also considers that set theory can be regarded as having an intuitive model in the cumulative hierarchy, but that does not make him commit to the existence of the entire set-theoretic universe, nor to accepting a Platonic stance:

\begin{adjustwidth}{1cm}{1cm}
\begin{quote}
\small
ZF has an intuitive model in the transfinite iteration of the power set operation taken cumulatively. This has nothing to do with a belief in a platonic reality whose members include the natural numbers and arbitrary sets of natural numbers, and so on. On the contrary, I disbelieve in such entities. \textcite[p. 72]{Feferman2000rationale}.
\end{quote}
\end{adjustwidth}

Perhaps Feferman is distinguishing between a non-standard, yet theoretically valid, model of ZF and a standard model of ZF, which would imply a further commitment to the reality of the universe it describes. He expressed a similar view in a later discussion of CH:

\begin{adjustwidth}{1cm}{1cm}
\begin{quote}
\small
There is no problem to put oneself in the mental frame of mind of “this is what the cumulative hierarchy looks like”, for which one can see that such and such propositions including the axioms of ZFC are (more or less) obviously true. I have taught set theory many times and have presented it in terms of this ideal-world picture with only the caveat that this is what things are supposed to be like in that world, rather than to assert that that's the way the world actually is. \textcite[p. 19]{Feferman2011a}.
\end{quote}
\end{adjustwidth}

Feferman's disbelief in a Platonic set-theoretical realm is again evident. His skepticism, however, does not seem to stem from a difficulty in conceiving the realm, but from the distance between such idealized structures and the real world. He concludes the paragraph with a comparison to idealized models in physics:

\begin{adjustwidth}{1cm}{1cm}
\begin{quote}
\small
I am no more uncomfortable doing that than if I were teaching Newtonian mechanics and talking of a world in which space-time is Euclidean (...), or if I were teaching relativistic mechanics. And one can see why little more is needed conceptually to develop an extraordinary part of set theory; as in the case of number theory, a little bit goes a long way. The issue is not whether we can think in such terms but, rather, what the philosophical status is of that kind of thought. \textcite[p. 19]{Feferman2011a}.
\end{quote}
\end{adjustwidth}

In summary, even though Feferman's requirement for clarity presents significant difficulties, it manages to capture important epistemological differences between arithmetic and set theory. However, extending these differences to third-order arithmetic and the minimal frameworks that allow us to formulate CH would be problematic. This is parallel to the conclusion of Section 4 regarding the applicability argument. In the next section, I will evaluate an important line of thought in Feferman's case on CH, which I will defend to be essentially a formal reformulation of his philosophical argumentation.


This is similar to the case around $CH$, where given a countable transitive model $M$ of $\mathrm{ZFC}$ we can force to obtain extensions $M[G_0]$, $M[G_0][G_1]$, $M[G_0][G_1][G_2]$,... that alternatively satisfy $CH$ and $\neg CH$. In fact, \textcite{Hamkins2012Multiverse} has used this fact as an argument that $CH$ should be regarded as indefinite.

Be $M \models \mathrm{PA}$ a model of Peano Arithmetic. An end-extension of $M$ is a larger model $N \supseteq M$ such that no element of $N - M$ is inserted ``bellow'' an old element of $M$, i.e., whenever $a \in M$ and $b \in N$ satisfy $b <^N a $, then $b \in M$.

If $\phi$ is an Orey sentence for $\mathrm{PA}$, then if $M$ is a model of $\mathrm{PA}$, we can end-extend it to obtain a model $N_1$ of $\mathrm{PA}$ where $\phi $ holds, and another model $N_2$ one where $\neg \phi$ does.


\section{Considerations around the clarity argument}
\label{clarity}

Feferman's central idea is that there is a progressive loss of clarity as we move from fully transparent concepts—such as that of a natural number sequence—to increasingly obscure notions:

\begin{adjustwidth}{1cm}{1cm}
\begin{quote}
\small
For example, we have a much clearer conception of arbitrary sequences of points on the Hilbert (or Dedekind, or Cauchy-Cantor) line, or at least of bounded strictly monotone sequence, than we do of arbitrary subsets of the line (...). Our conception allows us to reason that there is no enumeration of all infinite paths through the binary tree, resp. of all subsets of $N$, simply by the appropriate diagonal argument (...). And when we step to $S(S(N))$ there is a still further loss of clarity, but it is just the definiteness of that that is needed to make definite sense of CH. \textcite[p. 21]{Feferman2011a}.
\end{quote}
\end{adjustwidth}

Thesis 5 of conceptual structuralism—see Section 2—maintains that if a conception is not completely clear, then statements formulated within it cannot possess definite truth values. Thus, Feferman's argument would imply that CH is an indefinite statement, provided his reasons are compelling for regarding higher-order frameworks, such as third-order arithmetic or the cumulative set-theoretic hierarchy (the ones where CH can be precisely formalized), as lacking full clarity.


In addition to this, the idea of clarity of a determinate concept seems to be too subjective to constitute a criterion for its validity. This is something Koellner also points out:

\begin{adjustwidth}{1cm}{1cm}
\begin{quote}
\small
Intuitions of clarity are not robust. Some people think that the concept of an arbitrary subset of natural numbers is about as clear a conception as one can have; others disagree. Whose intuitions are to count? \textcite[p. 14]{Koellner2022}.
\end{quote}
\end{adjustwidth}

Feferman himself could be considered to lean towards interpreting clarity as an intrinsically subjective property of a concept, at least that seems to appear in Thesis 6 of conceptual structuralism (\textcite{Feferman2011a}, p. 11): “What is clear in a given conception is time-dependent, both for the individual and historically”. This makes Koellner's point even stronger.

Feferman's insistence on clarity can also be interpreted as a modern continuation of the predicative tradition, although he did not always explicitly identify with it:

\begin{adjustwidth}{1cm}{1cm}
\begin{quote}
\small
Though I am not now nor never have been a predicativist, I have to admit to being a sympathizer since I am an avowed anti-platonist, at least insofar as set theory is concerned, and I grant the natural numbers a position of primacy in our mathematical thought. \textcite[p. 313]{Feferman2004a}.
\end{quote}
\end{adjustwidth}

According to predicativism, mathematical entities should be conceptually well-defined to avoid paradoxes or ill-founded reasoning, though the required clarity is now framed in terms of formal, axiomatic structures rather than pre-theoretical intuition alone. Predicativism is hard to reconcile with the idea of defining the continuum set theoretically as a definite totality using the set $\mathcal{P}(\mathbb{N})$. This is essentially because, for a strict predicativist, the power set axiom implies accepting without a clear ground the set of all subsets of a given set, which is a totality that has meaning only in terms of its parts, not necessarily previously defined or built in any satisfactory way.

Let us recall the liar paradox, the statement that “This statement is false”. \textcite{Russell1908} rephrased it as follows: “It is not true for all propositions $p$ that if I affirm $p$, $p$ is true”. He then concluded that the cause of the paradox was a quantification over all the propositions, so that the notion of an arbitrary proposition is illegitimate. A similar reasoning was employed to explain the barber paradox, which appears after trying to define the naïve set ${x \mid x \notin x}$. This concept of an arbitrary naïve set was considered to be unclear, non-definite: this is what was thought to lead to the contradiction. Intuitionists tried to extend this debate even further, by stating that these paradoxes come from accepting logical structures that do not have a counterpart in our intuition (such as the law of excluded middle). We could perhaps say something similar about the concept of an arbitrary subset of the continuum: maybe this concept is so far from our intuition that trying to understand it in linguistic terms may lead us to contradictions or at least incongruities. However, there is no compelling evidence that the concept of an arbitrary subset of the continuum can lead to a contradiction in the same sense as the concept of a naïve set did, or the concept of all propositions. In fact, axiom systems such as ZF are hardly ever doubted to be consistent.


\section{Conclusion}

The Feferman-Koellner dispute on CH is just one more episode in the long line of philosophical and mathematical controversies that this conjecture has generated since Cantor posed it. Both Feferman and Koellner have made precise and insightful contributions to these controversies. In my view, the only significant weakness in Koellner's argument was his failure to consider Feferman's conceptual structuralism more fully. It is true that Feferman's writings are sometimes too concise and difficult to understand, but I believe the interpretation I have given here is loyal to his main ideas.

The fundamental thesis of this paper is that this philosophical conception, once interpreted as I do, offers a complementary argument for the indefiniteness of CH, which is grounded also in scientific and practical applicability, and not only in the notion of mathematical clarity, upon which Feferman explicitly relies. Considering the possibility of this applicability argument, I believe Koellner is incorrect in stating that conceptual structuralism does not address the indefiniteness of CH. However, this does not mean that we should accept the argument in its basic form, as presented in Section 3.

The next step should be to develop a more fine-grained notion of the applicability of mathematical structures, which includes both the notion of applicability for natural science and the notion of internal applicability to solve problems that arise within previously defined mathematical conceptions.

The class $S_0$ would consist of structures whose legitimacy is grounded directly in their indispensability in modeling the physical world, and would probably include structures such as the set of natural numbers $\mathbb{N}$, the rational numbers $\mathbb{Q}$, the Euclidean plane for spatial reasoning, and others representing core notions of real analysis and calculus (such as that of continuous change). Any structure in $S_0$ can then be justified through the external applicability requirement in its most direct form. Once $S_0$ is defined, it is the moment to take the recursive step. A structure $X$, not belonging to $S = S_1 \cup \dots \cup S_n$, is admitted into class $S_{n+1}$ if it meets, in the most standard, simple and natural way, one of the following conditions:

\begin{enumerate}
  \item \textbf{Problem-Solving Power:} $X$ provides solutions to previously existing, well-defined and relevant problems, i.e., to problems formulated within structures in $S$.

  \item \textbf{Explanatory Unification:} $X$ is able to reveal that previously seemingly disparate phenomena derived from structures in $S$ are in fact manifestations of a single underlying principle from $X$.

  \item \textbf{Generative Capacity:} $X$ naturally leads to new, fruitful lines of inquiry in $S$.
\end{enumerate}

This definition requires further explanation, but I do not have the space to provide it now. The main goal of \textcite{Blanco2} is precisely to further explain what I understand as standard, simple, and natural mathematical conceptions and to state more explicitly which structures could fit into each level of the $\{S_n\}$ hierarchy. In particular, it would shed considerable light on the debate around CH to understand where, within this hierarchy, both systems of third-order arithmetic and more abstract set-theoretic conceptions stand.

My purpose in this paper has been to show how the notion of practical and scientific applicability can play a crucial role in the debate over the definiteness of CH, and how this leads to the next step in understanding such an enigmatic and controversial statement.

%=========================



\printbibliography


\appendix

\section{Set-theoretical background}
\label{set-theory appendix}

This appendix briefly reviews some set-theoretical notions relevant to the discussion of the quasi-categoricity of second-order set theory given in Section \ref{categoricity-second-order-logic}.

Modern set theory commonly understands the universe of sets through the \emph{cumulative hierarchy}, also called the \emph{Von Neumann hierarchy}. The idea is that sets are generated in stages, where at each stage one forms new sets only from sets already constructed at previous stages. The hierarchy is defined recursively along the ordinals:

\[
V_0 = \varnothing,
\]

\[
V_{\alpha+1} = \mathcal{P}(V_\alpha),
\]

and for limit ordinals $\lambda$,

\[
V_\lambda = \bigcup_{\beta < \lambda} V_\beta.
\]

Here $\mathcal{P}(X)$ denotes the power set of $X$, that is, the set of all subsets of $X$. Intuitively, each stage collects together all sets that can be formed from earlier stages. The entire set-theoretic universe is then represented by

\[
V = \bigcup_{\alpha \in \mathrm{Ord}} V_\alpha,
\]

where $\mathrm{Ord}$ denotes the class of all ordinals.

The hierarchy proceeds through the ordinals. Ordinals generalize the natural numbers by extending the notion of order type beyond the finite.

Successor ordinals are ordinals of the form $\alpha+1$. At successor stages, new sets are generated by applying the power set operation to the previous stage. Limit ordinals, by contrast, have no immediate predecessor. Examples include $\omega$, the first infinite ordinal. At a limit stage $\lambda$, no genuinely new operation is performed; instead one simply takes the union of all earlier stages:

\[
V_\lambda = \bigcup_{\beta < \lambda} V_\beta.
\]

Thus the cumulative hierarchy alternates between power set formation at
successor stages and unions at limit stages.

A cardinal $\kappa$ is called \emph{regular} if it cannot be expressed as the union of fewer than $\kappa$ sets, each of cardinality smaller than $\kappa$. Equivalently, $\kappa$ is regular if there is no function

\[
f : \lambda \to \kappa
\]

with $\lambda < \kappa$ such that the image of $f$ is unbounded in $\kappa$. Intuitively, a regular cardinal cannot be ``built up gradually'' from a smaller number of smaller cardinals. The cardinal $\omega$ is regular, whereas $\aleph_\omega$ is singular.

A cardinal $\kappa$ is a \emph{strong limit cardinal} if for every
$\lambda < \kappa$,

\[
2^\lambda < \kappa.
\]

Equivalently, taking power sets of smaller cardinals never reaches $\kappa$. This notion is important because the cumulative hierarchy is generated by iterating the power set operation.

A cardinal $\kappa$ is called \emph{strongly inaccessible} if:

\begin{enumerate}
    \item $\kappa$ is uncountable,
    \item $\kappa$ is regular,
    \item $\kappa$ is a strong limit cardinal.
\end{enumerate}

Strongly inaccessible cardinals are important because the structure $V_\kappa$ behaves like a miniature universe of sets. In particular, $V_\kappa$ satisfies the axioms of ZFC. This is the key idea behind the quasi-categoricity theorem for second-order set theory.

Replacement states, roughly, that if one applies a definable function to a set, the resulting collection is again a set.

Formally, if $f$ is a definable function and $x$ is a set, then

\[
f(x) = \{f(t) : t \in x\}
\]

is also a set.

The regularity of $\kappa$ ensures that applying definable functions to sets already in $V_\kappa$ cannot produce sets of rank reaching or exceeding $\kappa$. Thus the outputs of such functions remain inside $V_\kappa$.

\end{document}

