%=========================
% Plantilla LaTeX Artículo Académico
% Compilar con: pdflatex + biber + pdflatex + pdflatex
%=========================
\documentclass[11pt,a4paper]{article}

%-------------------------
% Idioma y tipografía
%-------------------------
\usepackage[english]{babel}
\usepackage{csquotes}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

%-------------------------
% Márgenes y espaciado
%-------------------------
\usepackage[a4paper,margin=2.5cm]{geometry}
\usepackage{setspace}
\onehalfspacing

%-------------------------
% Matemáticas y símbolos
%-------------------------
\usepackage{amsmath,amssymb,amsthm}

%-------------------------
% Tablas y figuras
%-------------------------
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{changepage}

%-------------------------
% Enlaces y PDF
%-------------------------
\usepackage{hyperref}
\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  citecolor=blue,
  urlcolor=blue,
  pdftitle={The Continuum Hypothesis as a speech act},
  pdfauthor={Rodrigo Blanco Betes}
}

%-------------------------
% Listas, código y utilidades
%-------------------------
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{csquotes} % recomendado con biblatex

%-------------------------
% Bibliografía (biblatex + biber)
%-------------------------
\usepackage[
  backend=biber,
  style=apa,
  sortcites=true,
  url=true,
  doi=true
]{biblatex}
\DeclareLanguageMapping{english}{english-apa}
\addbibresource{refs.bib}

%-------------------------
% Metadatos del artículo
%-------------------------
\title{\textbf{Notes and references}}

\author{Rodrigo Blanco Betes}


\date{\today}

%-------------------------
% Entornos de teoremas (opcional)
%-------------------------
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}

\begin{document}
\maketitle
Does the formal independence of the Continuum Hypothesis from the axioms of Set Theory imply that the question has no meaning? Some, including Cohen himself, have argued that the answer to this question is “yes” [1]. But for this position to be credible either the other classical problems in set theory which have been similarly shown to be formally unsolvable would also have to be regarded as without meaning, or there must be an explanation as to why the problem of the Continuum Hypothesis is different. \textcite[p. 29]{woodin2004}

Should the axiom of Projective Determinacy be accepted as true? One argument that is frequently cited against this axiom is that there are interesting alternative theories for the projective sets, indeed there are some very difficult open questions about the possible theories for the projective sets in the context where Projective Determinacy is false. But this argument confuses two issues. Accepting Projective Determinacy as true does not deny the study of models in which it is false. For example, it is certainly the prevailing belief that the axioms for Number Theory are consistent. Nevertheless the study of models of Number Theory in which the axioms are inconsistent seems clearly an interesting program and a deep understanding of these models probably entails resolving the P = NP question of computational complexity. Do I believe the axiom of Projective Determinacy is true? This I find to be a rather frequently asked question of me. My answer is this. I believe the axiom of Projective Determinacy is as true as the axioms of Number Theory. \textcite[pp. 31-32]{woodin2004} 

For each such countable ordinal, $\alpha$, the structure
\[
\langle \mathcal{P}(\alpha), \alpha, \cdot, +, \in \rangle
\]
is reducible to the structure,
\[
\langle \mathcal{P}(\omega), \omega, \cdot, +, \in \rangle.
\]

Therefore I claim that the next natural structure is
\[
\langle \mathcal{P}(\omega_1), \omega_1, \cdot, +, \in \rangle,
\]
which is the standard structure for all sets of countable ordinals. It is well known that
this structure is equivalent to the structure of all sets of hereditary cardinality less than $\aleph_2$.

Whether or not CH holds is a first order property of this structure. Therefore any reasonably complete axiomatization for this structure should resolve CH. \textcite[p. 32]{woodin2004}

The existence of good sentences (which are $\Omega$-satisfiable) and the theorem that good sentences must imply that the Continuum Hypothesisis false would seem to narrow the possibilities for an argument that the problem of the Continuum Hypothesisis fundamentally different from the measure problem for the projective sets to essentially just the fact that neither $CH$ or ($\neg CH$) is $\Omega_{ZFC}$-valid. But accepting this as the reason that the problem of the Continuum Hypothesis is without meaning forces one to make the stronger claim that if a 2-sentence is not $\Omega_{ZFC}$-valid then the claim that it is true is not meaningful (noting that both $CH$ and $\neg CH$ are 2 sentences). However the plausibility of such a claim critically depends on the $\Omega$ Conjecture. \textcite[p. 44]{woodin2004}

In this quote from Woodin, a sentence $\phi$ is \textit{good} in case it is $\Omega$-satisfiable, i.e. if and only if there is a consistent theory $T$ such that $T \models_\Omega \phi$. This is equivalent to saying that, for any ordinal $\alpha$ and complete Boolean algebra $\mathbb{B}$, if $V^\mathbb{B}_\alpha \models T$, then $V^\mathbb{B}_\alpha \models \phi$. This means these sentences are true in a way that is absolute and stable under forcing.

The main idea of Woodin is that in case there are “enough” good sentences, then all these sentences should lead to the fact that $CH$ is false.

It has also been pointed out that the larger large cardinal axioms are supported by strong arguments of analogy. Typically, such arguments single out a combinatorial property of the first infinite cardinal $\omega$ and posit the existence of uncountable cardinals having an analogous property. This is usually justified by appealing to the uniform generation of sets in the cumulative hierarchy on the grounds that this uniformity should rule out the existence of ‘singularities’. It is fair to say that this line of reasoning is generally regarded as carrying less credence than justifications via reflection principles. \textcite[p. 97]{hauser2004}

Nevertheless acceptance of PD as an axiom was hampered by its lack of intrinsic evidence. That difficulty was overcome when it was shown that PD is implied by large cardinal axioms. More precisely, PD follows from the existence of infinitely many Woodin cardinals. \textcite[p. 99]{hauser2004}

Note: PD can be deduced by the existence of sufficiently large cardinal axioms. If we follow \textcite{Maddy1997} (MAXIMIZE and UNIFY), that would be an argument for accepting this axiom. However, I think that would be the case only in case fruitful conections of this axiom with more down-to-earth areas of mathematics are found. We have to consider to which extend the results on the proyective sets this axiom entails (Baire property, perfect set property, Lebesgue measurability) can be thus considered. Else, this area of research would be abbandoned through the years.

This also affords an improved understanding of the claim that PD is the ‘correct’ axiom for second order number theory. Not only does PD yield a complete structure theory for the projective sets, it is also implied by and in fact equivalent to many combinatorial principles which on the surface have nothing to do with determinacy. Most importantly, however, second order number theory as given by PD is shared by all sufficiently strong extensions of ZFC. The reason for this is the curious phenomenon that propositions implying the consistency of PD actually imply PD. \textcite[p. 100]{hauser2004}

When Hilbert revived Euclid’s ideas at the beginning of the twentieth century he declared this indispensability as the characteristic mark of axiomhood. In [29] [=\textcite{Hilbert1918}] for example, intrinsic plausibility is not even mentioned. (...) Hilbert’s account is inadequate in several respects. For one thing, he ignores the
principle that an axiom ought to have a higher degree of evidence than the theorems it proves. Another more subtle inadequacy is brought to light by projective determanacy: how could one possibly remain convinced of its consistency without believing it is true? \textcite[p. 102]{hauser2004}

The idea of the legitimacy hierarchy does essentially provide an extrinsic reason to accept axioms as true, i.e. whether they model the most suitable structures. So essentially I agree that axioms in mathematics are justified in the end by extrinsic reasoning. In regard to the last observation by Hauser, I think that a consistent statement in a determinate axiomatic system can be regarded as a false or meaningless mathematical statement, in any case not a true mathematical statement. That does not contradict how he goes on: "The higher epistemic weight of intrinsic plausibility appears to be a consequence of the fact that—in contrast with extrinsic evidence—there is at least one way in which it is directly linked with truth, namely when an axiom is ‘implied’ by the concept(s) it aims to capture". 

What we find in virtually all situations is that the opposing parties ultimately defend their position by invoking extra-mathematical elements in addition to mathematical facts. For example, opponents of the ‘axiom’ of constructibility argue that it runs counter to ‘the goals of set theory’ or that it violates ‘the intended meaning of the concept of set’. What these opaque phrases suggest is that the meanings of mathematical concepts reside in the relationship in which the mind stands to the objective reality of mathematics. Moreover, if the mind is assigned a constitutive role with respect to mathematics as a whole, that relationship lies beyond the explanatory reach of mathematical facts. \textcite[p. 103]{hauser2004}

It would be countersensical to exclude his subjective perspective from the investigation of the role
that evidence plays in the selection of axioms. Admittedly, this contrasts with the
concerns of ‘ordinary science’ where some standard of objectivity is presupposed and
subjectivity is regarded as a disturbing element. \textcite[p. 104]{hauser2004}

Husserl's signitive vs intuitive distintion is crutial in mathematics. The comprehension of these two kind of consciousness acts is what he calls \textit{fulfillment}. Relation to sense vs reference.

We experience how in intuition the same objectuality is intuitively realized which was “merely meant” in the symbolic act, and that it becomes intuitable exactly as something determined in such a way as it was initially merely thought (merely meant). \textcite[Part 2, §8]{husserl1913logische}




\printbibliography


\end{document}